---
title: "Base Plotting in **R**: Part III"
author: "Elie Gurarie"
subtitle: "[EFB 654: R and Reproducible Research](https://eligurarie.github.io/EFB654/)"
date: "**February 18, 2026**"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", dpi = 200, cache = TRUE)
```

```{r load-data, include=FALSE, eval = FALSE, echo = FALSE}
library(WDI)

wdi_raw <- WDI(
  indicator = c(
    LifeExp   = "SP.DYN.LE00.IN",   # life expectancy at birth
    GDP       = "NY.GDP.PCAP.CD",   # GDP per capita (current USD)
    Area      = "AG.LND.TOTL.K2",   # total land area (sq km)
    Population= "SP.POP.TOTL"       # total population
  ),
  start = 2020, end = 2022,
  extra = TRUE   # pulls region, capital, longitude, latitude, income level
)

library(dplyr)

wdi <- wdi_raw |>
  filter(!is.na(GDP), !is.na(LifeExp)) |>
  group_by(country) |>
  slice_max(year, n = 1) |>
  ungroup() |>
  filter(region != "Aggregates")   # drops World Bank regional summaries

wdi <- subset(wdi, LifeExp > 40)

write.csv(wdi, file = "WDI.csv", row.names = FALSE)
```

# Goal: 

- Layouts (`mfrow` and `layout`)
- Exporting figures (`png`, `pdf`, `jpg` and in Makdown)
- Brief intro to `ggplot`


Using the World Development Indicators data from the [last lab](PlottingPartII_lab.html)

```{r}
wdi <- read.csv("WDI.csv")
wdi$region <- factor(wdi$region)
cols <- c("East Asia & Pacific"        = "steelblue",
          "Europe & Central Asia"      = "forestgreen",
          "Latin America & Caribbean"  = "goldenrod",
          "Middle East & North Africa" = "darkorange",
          "North America"              = "mediumpurple",
          "South Asia"                 = "firebrick",
          "Sub-Saharan Africa"         = "sienna")
```

Recall: 

```{r}
str(wdi)
```

# 1. Controlloing layouts

## with `mfrow()`

```{r, fig.width = 8, fig.height = 6}
par(mfrow = c(2,1))
boxplot(GDP/1000 ~ region, data = wdi, log = "y", las = 1, col = cols)
boxplot(LifeExp ~ region, data = wdi, las = 1, col = cols)
```

## Or prettier: modifying morgins with `oma` and `outer`

```{r, fig.width = 12, fig.height = 8}
par(mfrow = c(2,1), 
    mar = c(0,4,1,2), # margins around each figure
    oma = c(3,0,2,0), # outer margins
    cex.axis = 0.8, 
    tck = 0.01, bty ="l", mgp = c(2,.25,0))

boxplot(GDP/1000 ~ region, data = wdi, log = "y", las = 1, 
        col = cols, 
        xlab = "", xaxt = "n")
boxplot(LifeExp ~ region, data = wdi, las = 1, col = cols)
title("Global Indicators across regions", outer = TRUE)
```


## Issues with labelling? 

```{r, fig.width = 10, fig.height = 8}
par(mfrow = c(2,1), 
    mar = c(0,4,1,2), # margins around each figure
    oma = c(3,0,2,0), # outer margins
    cex.axis = 0.8, 
    tck = 0.01, bty ="l", mgp = c(2,.25,0))

boxplot(GDP/1000 ~ region, data = wdi, log = "y", las = 1, col = cols, 
        varwidth = TRUE, 
        xlab = "", xaxt = "n")  # no axis labelling
boxplot(LifeExp ~ region, data = wdi, las = 1, 
        varwidth = TRUE, 
        col = cols, xaxt = "n")
title("Global Indicators across regions", outer = TRUE)
regions <- levels(factor(wdi$region))
region_labels <- gsub("&", "&\n", regions, fixed = TRUE)
mtext(region_labels, side = 1, at = 1:7, line = 0, padj = 1)
```


## More complicated layouts with `layout`

```{r layout}
layout(rbind(c(1,2), c(1,3)))
layout.show(3)
```

## Now do this:

```{r FunLayout2, fig.width = 12, fig.height = 6}
require(scales)   # for transparency!
layout(rbind(c(1,2), c(1,3)))
par(cex.lab = 1.2, las = 1,
    mgp = c(2, .25, 0), tck = 0.01, 
    bty = "l",  mar = c(0,4,0,2), 
    oma = c(4,0,4,0), xpd = NA)

cont_col <- cols[match(wdi$region, levels(wdi$region))]
cex_pop  <- sqrt(wdi$Population / max(wdi$Population, na.rm = TRUE)) * 15

plot(wdi$GDP/1e3, wdi$LifeExp, log = "x",
     ylab = "Life expectancy (years)",
     xlab = "GDP per capita (1000 USD, log scale)", type = "n")
points(wdi$GDP/1e3, wdi$LifeExp,
       cex = cex_pop + .5, pch = 21, col = cont_col,
       bg  = alpha(cont_col, .7))
legend("bottomright", title = "Region",
       legend = levels(wdi$region),
       pt.bg = cols, col = "grey40",
       pch = 21, pt.cex = 1.8, bty = "n")

boxplot(GDP/1000 ~ region, data = wdi, log = "y", las = 1, 
        varwidth = TRUE, 
        col = alpha(cols, .7), border = cols, 
        xlab = "", xaxt = "n")  # no axis labelling
boxplot(LifeExp ~ region, data = wdi, las = 1, 
        varwidth = TRUE, 
        col = alpha(cols, .7), border = cols, 
        xaxt = "n", xlab = "Region")
title("Global Indicators across regions", outer = TRUE)
```


# 1b. Brief (unexpected) aside on functions

We'll talk about functions more later in class.  But one very basica pplication for using functions is bundling a bunch of code into a single quick line.  

For example:

```{r}
setPars <- function(){
  par(cex.lab = 1.2, las = 1,
    mgp = c(2, .25, 0), tck = 0.01, 
    bty = "l",  mar = c(0,4,0,2), 
    oma = c(4,0,4,0), xpd = NA)
}


plotWDI <- function(){
  cont_col <- cols[match(wdi$region, levels(wdi$region))]
  cex_pop  <- sqrt(wdi$Population / max(wdi$Population, na.rm = TRUE)) * 15
  
  plot(wdi$GDP/1e3, wdi$LifeExp, log = "x",
       ylab = "Life expectancy (years)",
       xlab = "GDP per capita (1000 USD, log scale)", type = "n")
  points(wdi$GDP/1e3, wdi$LifeExp,
         cex = cex_pop + .5, pch = 21, col = cont_col,
         bg  = alpha(cont_col, .7))
  legend("bottomright", title = "Region",
         legend = levels(wdi$region),
         pt.bg = cols, col = "grey40",
         pch = 21, pt.cex = 1.8, bty = "n")
}

boxplotWDI <- function(){
  boxplot(GDP/1000 ~ region, data = wdi, log = "y", las = 1, 
        varwidth = TRUE, 
        col = alpha(cols, .7), border = cols, 
        xlab = "", xaxt = "n")  # no axis labelling
  boxplot(LifeExp ~ region, data = wdi, las = 1, 
          varwidth = TRUE, 
          col = alpha(cols, .7), border = cols, 
          xaxt = "n", xlab = "Region")
}
```

Note - these functions ONLY do the plotting - none of the setup.  Also they have no arguments and they return nothing.  They just do the thing.  Code then becomes very simple:

```{r PlottingSuperQuick, fig.width = 12, fig.height = 6}
layout(rbind(c(1,2), c(1,3)))
setPars()
plotWDI()
boxplotWDI()
```



# 2. Outputting figures to files

## What is a graphics device?

When R draws a plot, it sends output to a **graphics device** — a destination that knows how to render lines, points, text, and colors. By default that destination is your screen (the interactive device: `windows()` on Windows, `quartz()` on Mac). To save a figure to a file, you open a **file device** instead, draw your plot, then close the device with `dev.off()`. The pattern is always the same:

```{r device-pattern, eval=FALSE}
device_function("filename.ext", width = ..., height = ...)  # open
  # ... plotting code ...
dev.off()                                                    # close and write
```

Nothing appears on screen — output goes straight to the file. Forgetting `dev.off()` is the most common mistake: the file stays open and incomplete.

---

## PDF vs. PNG: vector vs. raster

The two most useful formats are **PDF** (vector) and **PNG** (raster). They store images differently and suit different purposes.

| | PDF | PNG |
|---|---|---|
| Type | Vector | Raster (bitmap) |
| Stores images as | Mathematical shape descriptions | Grid of pixels |
| Scaling | Perfect at any size | Degrades if enlarged |
| Best for | Publications, LaTeX, print | Web, Word, presentations |
| Resolution setting | Not applicable | Critical — use `res =` |

**PDF** is resolution-independent: a line is stored as "from point A to point B," so it renders crisp at any zoom level or print size. Ideal for journal submissions.

**PNG** stores a fixed pixel grid. At low resolution it looks blurry in print; at high resolution files get large. The `res` argument (dots per inch) controls the tradeoff.

---

## Exporting to PDF

```{r pdf-example, eval=FALSE}
pdf("WDImegaplot.pdf", width = 12, height = 6)   # dimensions in inches

  layout(rbind(c(1,2), c(1,3)))
  setPars()
  plotWDI()
  boxplotWDI()

dev.off()
```

- `width` and `height` are always **inches**
- Default is `7 × 7` inches
- Plotting in a loop produces a **multi-page PDF** — one page per plot

---

## Exporting to PNG

```{r png-example, eval=FALSE}
png("WDImegaplot.png", width = 2400, height = 1200, res = 200)   # pixels

  layout(rbind(c(1,2), c(1,3)))
  setPars()
  plotWDI()
  boxplotWDI()
  
dev.off()
```

- `width` and `height` are **pixels** by default
- `res` sets dots per inch — controls how large text and points appear relative to the canvas
- Effective plot size in inches = `width / res`, so `width = 800, res = 150` gives a ~5.3 × 4 inch plot
- For print quality use `res = 300`; for screen `res = 96` or `res = 150` is fine

---

## Controlling size in R Markdown

In an `.Rmd` file you don't open and close devices manually — knitr handles that. Instead, control figure size through **chunk options**:

````
```{r myplot, fig.width = 8, fig.height = 5, dpi = 200}
# ... plotting code ...
```
````

- `fig.width` / `fig.height` — dimensions in **inches** (like PDF)
- `dpi` — resolution of the rendered PNG (knitr defaults to 72; use 150–300 for print)
- Set defaults for the whole document in the setup chunk:

```{r setup-example, eval=FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 5, dpi = 150)
```


---

## `jpeg()` — when to use it

`jpeg()` works identically to `png()` but uses lossy compression:

```{r jpeg-example, eval=FALSE}
jpeg("myplot.jpg", width = 800, height = 600, res = 150, quality = 90)
  # ... plotting code ...
dev.off()
```

- `quality` ranges 0–100; lower = smaller file, more compression artifacts
- Avoid for plots with sharp lines, text, or flat color regions — artifacts are visible
- Reasonable for photographic content or where file size is the priority




# 3. Introduction to `ggplot2`

## History and philosophy

`ggplot2` was created by Hadley Wickham in 2005, based on Leland Wilkinson's book *The Grammar of Graphics* (1999) — hence the "gg". The core idea is that every statistical graphic can be described by a small set of components assembled in a consistent way:

- **data** — a data frame
- **aesthetics** — mappings from variables to visual properties (x, y, color, size, ...)
- **geometries** — the actual marks drawn (points, lines, bars, ...)
- **scales**, **facets**, **themes** — further refinements

In base R, you build a plot by issuing a sequence of imperative commands (`plot()`, then `points()`, then `legend()`, ...). In `ggplot2`, you *declare* what you want and the package figures out the rendering. The result is a single object you can inspect, modify, and print.

---

## Everything is built from data frames

This is the single most important constraint: **every variable you want to plot must be a column in a data frame**. You cannot pass a loose vector to `ggplot2` the way you can to `plot()`. This is sometimes inconvenient, but it forces a discipline that pays off with complex data.

```{r ggplot-load, message=FALSE}
library(ggplot2)
wdi <- read.csv("WDI.csv")
wdi$region <- factor(wdi$region)
```

---

## Aesthetics and geoms

A `ggplot2` call has two required ingredients: `ggplot()` sets up the data and aesthetic mappings; a `geom_*()` function specifies what to draw.

```{r gg-basic, fig.width=7, fig.height=5}
ggplot(data = wdi, aes(x = GDP/1e3, y = LifeExp)) +
  geom_point()
```

`aes()` maps **variables** to **visual channels**:

| aesthetic | visual property |
|-----------|----------------|
| `x`, `y`  | position |
| `color`   | point/line color (outline) |
| `fill`    | filled shape color |
| `size`    | point size |
| `alpha`   | transparency |
| `shape`   | point character |

Common `geom_*` functions:

| geom | what it draws |
|------|--------------|
| `geom_point()` | scatter plot |
| `geom_line()` | lines connecting observations |
| `geom_smooth()` | smoothed trend with confidence band |
| `geom_boxplot()` | box-and-whisker |
| `geom_histogram()` | histogram |
| `geom_bar()` / `geom_col()` | bar chart |
| `geom_text()` | text labels |

---

## Advantage 1 — adding dimensions is quick

In base R, encoding a third or fourth variable requires computing color vectors, size scalings, and legend calls by hand. In `ggplot2`, you just add aesthetics — scales and legends are generated automatically.

Compare adding color, size, and a log scale:

```{r gg-multidim, fig.width=9, fig.height=6}
ggplot(wdi, aes(x = GDP/1e3, y = LifeExp, color = region, size  = Population/1e6)) +
  geom_point(alpha = 0.7) +
  scale_x_log10() +
  theme_bw()
```

Four variables (GDP, life expectancy, region, population) in about 4 lines, with automatic legends for both color and size.


---

## BUT --- it is extremely difficult to customize

`ggplot2`'s layered system makes common tasks easy but uncommon tasks surprisingly hard. Moving a legend to an exact coordinate, controlling tick positions on a secondary axis, mixing fonts within a plot title, nudging individual labels — all of these require either digging into `theme()` (which has ~100 arguments) or reaching for extension packages like `ggtext`, `ggnewscale`, or `ggrepel`.

In base R, arbitrary customization is usually a few `par()` arguments or a call to `text()` / `mtext()`. The same result in `ggplot2` may require a stack of overrides.

```{r gg-theme-pain, eval=FALSE}
# Moving legend inside the plot, changing its background,
# adjusting axis text size, removing the minor grid,
# and making the title left-aligned — all separate theme() calls
theme(
  legend.position      = c(0.15, 0.85),
  legend.background    = element_rect(fill = "white", color = "grey70"),
  axis.text            = element_text(size = 11),
  panel.grid.minor     = element_blank(),
  plot.title.position  = "plot"
)
```

> Customizing a ggplot is one of the better uses of Large-Language Models

---

## Faceting is extremely powerful / easy for data exploration

Faceting splits the data into subplots by one or more categorical variables — something that requires manual `layout()` calls and repeated plotting code in base R. In `ggplot2` it is a single line.

```{r gg-facet, fig.width=11, fig.height=8}
ggplot(wdi, aes(x = GDP/1e3, y = LifeExp, size = Population/1e6)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", color = "firebrick",linewidth = 0.7) +
  scale_x_log10() +
  facet_wrap(~ region)   # key line of code
```

`facet_wrap(~ region)` produces one panel per region, all on the same scales, with a trend line in each. Doing this in base R would require a loop, careful margin management, and separate legend placement.  

`facet_grid(rows ~ cols)` extends this to two categorical variables simultaneously.

The equivalent task in base R plotting is much harder.  

---

## ggplot2 vs. base R: a summary

| | base R | ggplot2 |
|---|---|---|
| Learning curve | Steeper for basics | Steeper for customization |
| Multi-variable plots | Manual (but explicit) | Fast (automatic legends/scales) |
| Faceting | Manual `layout()` | `facet_wrap()` / `facet_grid()` |
| Fine-grained control | Excellent | Difficult |
| Default aesthetics | Minimal, print-friendly | Ugly (IMO) |
| Data requirement | Any form of data is OK | Data frame required |

The two approaches are complementary. `ggplot2` is often faster for exploratory work and for panel figures; base R is often better when you need precise, publication-ready control over every element.
