---
title: "Homework 1: Getting Started with R and RMarkdown"
author: "Your Name Here"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions

Complete all exercises below by filling in your answers in the designated spaces. When finished:

1. Save this .Rmd file
2. Knit the document to HTML (click the "Knit" button or press Ctrl+Shift+K / Cmd+Shift+K)
3. Submit **both** your .Rmd file and the knitted Word/PDF (HTMLs are not 'uploadable', but can be converted to PDF) to Blackboard

**Due:** Wednesday, January 21 by 11:59 PM

---

## Part 1: About You

### Question 1 (1 point)

Tell us about yourself! In the space below, write a few paragraphs covering:

- Your name and program (undergraduate major or graduate department)
- Your research interests or thesis/dissertation topic (if applicable)
- Your prior experience with programming or statistical software (R, Python, Excel, etc.)
- What you hope to get out of this course

*Delete this text and write your response here.*

---

### Question 2 (0.5 points)

Then, in a sentence, tell us what subject area and packages within TaskViews you're most interested in.

*Delete this text and write your response here.*

---

## Part 2: Basic R Operators

For each question below, write R code in the provided code chunk and interpret the results.

### Question 3 (1 points)

Use R as a calculator to compute the following. Show your code and the output.

a) $25 + 17$
b) $144 / 12$
c) $8^3$ (8 cubed)
d) The square root of 256 (hint: use `sqrt()` or the exponent 0.5)

```{r q3}
# Write your code here

```

---

### Question 4 (1 point)

R has several comparison operators that return `TRUE` or `FALSE`. Predict what each of the following will return, then run the code to check your prediction.

a) `5 > 3`
b) `10 == 11`
c) `7 != 7`
d) `15 >= 15`

```{r q4}
# Write your code here

```

In your own words, explain what the `==` and `!=` operators do:

*Delete this text and write your explanation here.*

---

### Question 5 (0.5 points)

Create three variables:

- `x` assigned the value 10
- `y` assigned the value 3
- `z` assigned the result of `x` times `y`

Then print the value of `z`.

```{r q5}
# Write your code here

```

---

### Question 6 (1 point)

The modulo operator `%%` returns the remainder after division. For example, `10 %% 3` returns `1` because 10 divided by 3 is 3 with a remainder of 1. 

a) What does `17 %% 5` return? Run the code and explain why.
b) What does `20 %% 4` return? Explain why.
c) How could you use the modulo operator to determine if a number is even or odd? Demonstrate with an example.

```{r q6}
# Write your code here

```

*Delete this text and write your explanations here.*

---

### Question 7 (1 point)

The `&` (AND) operator returns `TRUE` only if *both* conditions are true. The `|` (OR) operator returns `TRUE` if *either* condition is true.

Given `a <- 10` and `b <- 20`, predict what each expression will return, then run the code to check:

a) `a > 5 & b > 25`
b) `a > 5 | b > 25`
c) `a == 10 & b == 20`

```{r q7}
# Write your code here

```

In your own words, explain the difference between `&` and `|`:

*Delete this text and write your explanation here.*

---

## Submission Checklist

Before submitting, make sure you have:

- [ ] Filled in your name in the author field at the top
- [ ] Answered all questions
- [ ] Successfully knitted your document to HTML (or Word/PDF)
- [ ] Uploaded both your .Rmd file AND your knitted output to Blackboard

---

*This homework is worth 5.5 points total.*
