---
title: "Homework 8: Apply and ShinyApps"
author: 
date: "Assigned Apr 6, 2026; Due Apr 15, 2026"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
**Due:** Wednesday, April 15 by 11:59 PM

Complete the exercises below.  

- For Part I - create an `.Rmd` file with the solutions, knit it to html or pdf, and upload both

- For Part II - create and a separate script (e.g. `faithful_visualization.R`) with the working Shiny app OR, for extra credit, upload & deploy the app to `shinyapps.io`.  


###  I. Practice being a list-ninja (**4pts**)

Referring to [this lecture](https://eligurarie.github.io/EFB654/lectures/17-apply-and-lists/apply-mentality-slides.html], especially the exercises [here](https://eligurarie.github.io/EFB654/lectures/17-apply-and-lists/apply-mentality-slides.html#19):


1. Use `sapply` to compute the number of NA values in each column of the `airquality` dataset preloaded in R. *Hint: combine `is.na()` and `sum()`.*

2. Use `sapply` to extract the residual standard error from each model in fits. *Hint: this lives in  `summary(fit)$sigma`*.

3. Use `split` + `sapply` to compute the maximum Ozone value for each month, handling NA.

4. Perform the last operation in one line using `plyr::ddply()` and `summarize` OR `dplyr::group_by` and `summarize`.  


### Practice with Shiny (**4pts**)

See the [Shiny App lecture](https://eligurarie.github.io/EFB654/lectures/19-shinyapps/shiny-lab.html) The code for the Old Faithful Geyser visualization app is below (click on `Show`).  

Add a new tab to this app that:
    a. Plotsthe relationship between eruption duration and waiting times
    b. Prints some summary statistics on that relationship, notably the intercept, slope, *p*-value of linear model and R^2^ value.  
    ***Hint***  *as above, these are all located in the output of `summary(lm(eruptions~waiting, data = faithful))`.*  



```{r, eval = FALSE}
library(shiny)

ui <- fluidPage(
  titlePanel("Old Faithful Geyser Data"),
  sidebarLayout(
    sidebarPanel(
      sliderInput("bins", "Number of bins:", min = 1, max = 50, value = 30),
      selectInput("color", "Bar color:",
                  choices = c("darkgray", "steelblue", "tomato",
                              "forestgreen", "goldenrod"),
                  selected = "darkgray")
    ),
    mainPanel(
      tabsetPanel(
        tabPanel("Waiting Time",
                 plotOutput("waitPlot"),
                 verbatimTextOutput("waitStats")),
        tabPanel("Eruption Duration",
                 plotOutput("eruptPlot"),
                 verbatimTextOutput("eruptStats"))
      )
    )
  )
)

server <- function(input, output) {

  output$waitPlot <- renderPlot({
    x <- faithful$waiting
    bins <- seq(min(x), max(x), length.out = input$bins + 1)
    hist(x, breaks = bins, col = input$color, border = "white",
         xlab = "Waiting time (mins)", main = "Waiting Time")
  })

  output$eruptPlot <- renderPlot({
    x <- faithful$eruptions
    bins <- seq(min(x), max(x), length.out = input$bins + 1)
    hist(x, breaks = bins, col = input$color, border = "white",
         xlab = "Eruption duration (mins)", main = "Eruption Duration")
  })

  output$waitStats  <- renderPrint({ summary(faithful$waiting) })
  output$eruptStats <- renderPrint({ summary(faithful$eruptions) })
}

shinyApp(ui = ui, server = server)
```


    
**Extra Credit:**  Set up a free `shinyapps.io` account, add your name to the app somewhere (e.g. in the title), and share the link.  