---
title: Importing and Exporting data
date: Feb 9, 2026
output: 
  ioslides_presentation:
    fig_width: 7
    fig_height: 6
    fig_caption: true
    widescreen: true
    css: custom.css
---

Today
----
 
 - `read.table()` and subfunctions
 - `write.table()` and `write.csv()`
 - `readLines()`
 - `writeLines()`
 - more about `save()`
 
`read.table()`
----

A lot of options...
<PRE>
read.table(file, header = FALSE, sep = "", quote = "\"'",
           dec = ".", numerals = c("allow.loss", "warn.loss", "no.loss"),
           row.names, col.names, as.is = !stringsAsFactors,
           na.strings = "NA", colClasses = NA, nrows = -1,
           skip = 0, check.names = TRUE, fill = !blank.lines.skip,
           strip.white = FALSE, blank.lines.skip = TRUE,
           comment.char = "#",
           allowEscapes = FALSE, flush = FALSE,
           stringsAsFactors = FALSE,
           fileEncoding = "", encoding = "unknown", text, skipNul = FALSE)
</PRE>

 - **`file`** = file name you want to read in as a character (in quotes)
 - **`header`** = does it have a header? (T or F)
 - **`sep`** = what symbol is separating the values in each column

`read.table()`
----

...more options...
<PRE>
read.table(file, header = FALSE, sep = "", quote = "\"'",
           dec = ".", numerals = c("allow.loss", "warn.loss", "no.loss"),
           row.names, col.names, as.is = !stringsAsFactors,
           na.strings = "NA", colClasses = NA, nrows = -1,
           skip = 0, check.names = TRUE, fill = !blank.lines.skip,
           strip.white = FALSE, blank.lines.skip = TRUE,
           comment.char = "#",
           allowEscapes = FALSE, flush = FALSE,
           stringsAsFactors = FALSE,
           fileEncoding = "", encoding = "unknown", text, skipNul = FALSE)
</PRE>

 - **`quote`** = What characters are used to denote quotes?
 - **`dec`** = What's used to denote decimals? Usually "." or ",".
 - **`row.names`** = which column are the rownames in?
 - **`col.names`** = specify column names directly
 
`read.table()`
----

...more options...
<PRE>
read.table(file, header = FALSE, sep = "", quote = "\"'",
           dec = ".", numerals = c("allow.loss", "warn.loss", "no.loss"),
           row.names, col.names, as.is = !stringsAsFactors,
           na.strings = "NA", colClasses = NA, nrows = -1,
           skip = 0, check.names = TRUE, fill = !blank.lines.skip,
           strip.white = FALSE, blank.lines.skip = TRUE,
           comment.char = "#",
           allowEscapes = FALSE, flush = FALSE,
           stringsAsFactors = FALSE,
           fileEncoding = "", encoding = "unknown", text, skipNul = FALSE)
</PRE>

 - **`as.is`** = Which columns to keep strings as factors (prevent conversion)?
 - **`na.strings`** = What character symbols should be interpreted as NAs?
 - **`colClasses`** = A vector of classes to be assumed for columns
 
`read.table()`
----

...more options...
<PRE>
read.table(file, header = FALSE, sep = "", quote = "\"'",
           dec = ".", numerals = c("allow.loss", "warn.loss", "no.loss"),
           row.names, col.names, as.is = !stringsAsFactors,
           na.strings = "NA", colClasses = NA, nrows = -1,
           skip = 0, check.names = TRUE, fill = !blank.lines.skip,
           strip.white = FALSE, blank.lines.skip = TRUE,
           comment.char = "#",
           allowEscapes = FALSE, flush = FALSE,
           stringsAsFactors = FALSE,
           fileEncoding = "", encoding = "unknown", text, skipNul = FALSE)
</PRE>

 - **`nrows`** = How many total rows to read in (after the header, if applicable)
 - **`skip`** = How many rows to skip before start reading. This includes skipping the header.
 - **`check.names`** = Makes sure names are valid column names. No special characters, etc.
 
`read.table()`
----

...more options...
<PRE>
read.table(file, header = FALSE, sep = "", quote = "\"'",
           dec = ".", numerals = c("allow.loss", "warn.loss", "no.loss"),
           row.names, col.names, as.is = !stringsAsFactors,
           na.strings = "NA", colClasses = NA, nrows = -1,
           skip = 0, check.names = TRUE, fill = !blank.lines.skip,
           strip.white = FALSE, blank.lines.skip = TRUE,
           comment.char = "#",
           allowEscapes = FALSE, flush = FALSE,
           stringsAsFactors = FALSE,
           fileEncoding = "", encoding = "unknown", text, skipNul = FALSE)
</PRE>

 - **`fill`** = Should missing values be filled in?
 - **`strip.white`** = Should spaces or tabs before or after values be removed?
 - **`blank.lines.skip`** = Should blank rows be ignored?
 
Know the options for `read.table()`
----

If you know how they work, you will save time by avoiding typing, find-and-replacing, etc.

`read.csv()` and `read.delim()`
----

".csv" = comma-separated values

<PRE>
read.csv(file, header = TRUE, sep = ",", quote = "\"",
         dec = ".", fill = TRUE, comment.char = "", ...)
</PRE>
Tab-delimited values

<PRE>
read.delim(file, header = TRUE, sep = "\t", quote = "\"",
           dec = ".", fill = TRUE, comment.char = "", ...)
</PRE>

Wrappers for `read.table()`
----


Access the *object* `read.csv`.
```{r}
read.csv
```


Read Puromycin data
----

List all files in the directory with the extension ".csv"
```{r}
list.files(pattern=".csv")
```
 
"The Puromycin data frame has 23 rows and 3 columns of the reaction velocity versus substrate concentration in an enzymatic reaction involving untreated cells or cells treated with Puromycin."---R help
 
Written as a ".csv" file

<PRE>
"","conc","rate","state"
"1",0.02,76,"treated"
"2",0.02,47,"treated"
"3",0.06,97,"treated"
"4",0.06,107,"treated"
"5",0.11,123,"treated"
"6",0.11,139,"treated"
"7",0.22,159,"treated"
"8",0.22,152,"treated"
"9",0.56,191,"treated"
"10",0.56,201,"treated"
"11",1.1,207,"treated"
"12",1.1,200,"treated"
"13",0.02,67,"untreated"
"14",0.02,51,"untreated"
"15",0.06,84,"untreated"
"16",0.06,86,"untreated"
"17",0.11,98,"untreated"
"18",0.11,115,"untreated"
"19",0.22,131,"untreated"
"20",0.22,124,"untreated"
"21",0.56,144,"untreated"
"22",0.56,158,"untreated"
"23",1.1,160,"untreated"
</PRE>

Read Puromycin data
----

```{r}
pur<-read.csv("Puromycin_data.csv", row.names=1)
head(pur)
```

Write Puromycin data 
----

Change to *tab-delimted text* using `write.table()`

```{r}
write.table(pur, file = "Puromycin_data.delim", row.names=F, sep= "\t")
```

 - There is no `write.delim()` function in base R.

<PRE>
"conc"  "rate"	"state"
0.02	76	"treated"
0.02	47	"treated"
0.06	97	"treated"
0.06	107	"treated"
0.11	123	"treated"
0.11	139	"treated"
0.22	159	"treated"
0.22	152	"treated"
0.56	191	"treated"
0.56	201	"treated"
1.1	207	"treated"
1.1	200	"treated"
0.02	67	"untreated"
0.02	51	"untreated"
0.06	84	"untreated"
0.06	86	"untreated"
0.11	98	"untreated"
0.11	115	"untreated"
0.22	131	"untreated"
0.22	124	"untreated"
0.56	144	"untreated"
0.56	158	"untreated"
1.1	160	"untreated"
</PRE>

`readLines()` 
----

What if we just downloaded a massive file with an unknown extension, but we think it's text.

```{r}
file.info("usearch.otus.97.tax")
```

`readLines()`
----

```{r}
tax<-readLines("usearch.otus.97.tax", n=4)
head(tax)
```

 - `readLines()` can be used to quickly sample and explore unfamiliar data

`cat()` and `writeLines()`
----

 - `cat()` = raw flexible output to file or console

```{r}
month.abb[1:4]
```

```{r}
cat(month.abb[1:4])
```

`cat()` and `writeLines()`
----

 - `writeLines()` = similar to `cat()`, but default `sep = "\n"`
```{r}
month.abb[1:4]
```
```{r}
writeLines(month.abb[1:4])
```

 - `cat()` is probably not what you want unless you specify `sep`.

Saving big files
----

Read in "tax" file:
```{r, results='hold'}
system.time(tax<-read.delim("usearch.otus.97.tax"))
dim(tax)
object.size(tax)
```

Save as R object with compression
----

```{r, eval=FALSE}
save(tax, file = "TAX_data_GZIP.Rdata", compress="gzip")
```

```{r}
list.files()
```

Compressed vs Original
----

Compressed R object 10X smaller than original text file
```{r, results='hold'}
file.info("usearch.otus.97.tax")[1] / 10^6 #Original text file
file.info("TAX_data_GZIP.Rdata")[1]/ 10^6 # Compressed R object
```
 
 - Multiple compression methods
 - Compression adds time for packing and unpacking

Summary
----

 - Ensure columns have a single class of data only, i.e., either numeric, factor, or character. Get rid of values like "<5".
 - Know the options of `read.table()` and related functions.
 - Windows users: check out packages `RExcel`, `xlsx`, `XLConnect`, or `gdata`.

Try it Yourself
----

 1. Import the `qPCR-data.txt` file into `R` removing the metadata at the top of the file. 
 2. Find the median "CT" of all wells.
 3. Save the data frame as a compressed R object.
 
Try it Yourself
----

1. Import the `qPCR-data.txt` file into `R` removing the metadata at the top of the file. 

I did it this way:
```{r eval=T}
dat<-read.delim("qPCR-data.txt", skip=41, na.strings="Undetermined")
```

Try it Yourself {.smaller}
----

2.Find the median "CT" of all wells.     

```{r eval=T}
head(dat[, 1:10])
median(dat$CT, na.rm=T)
```

----

3.Save the data frame as a compressed R object.

```{r eval=T}
save(dat, file="qPCR-data.Rdata", compress = "gzip")
list.files()
```
