---
title: Indexing Data Frames
date: Feb 04, 2026
output: 
  ioslides_presentation:
    fig_width: 7
    fig_height: 6
    fig_caption: true
    widescreen: true
    keep_md: true
---

```{r echo=F}
knitr::opts_knit$set(keep.md = TRUE)
```

Short Refresher on Indexing Vectors {.smaller}
----

```{r}
data(islands)
length(islands)
islands
```

Index a vector with its **INDEX** (always an **INTEGER**)
----

```{r}
islands[3]
islands[3:6]
islands[-c(1,2)] #REMOVE values "Africa" and "Antarctica"
```

Index a vector by **NAME**
----

```{r}
islands["Asia"]
islands[c("Asia", "Java")]
```

Index a vector with **LOGICALS**
----

First create a logical vector
```{r}
log.vec<-islands > 100 & islands < 1000 # TRUE for islands > 100 & less than 1000
str(log.vec)
log.vec
```

Index a vector with **LOGICALS**
----

```{r}
islands[log.vec]
```

Also remember **"Recycling"**
----

```{r}
str(islands) # islands is 48 length
islands[c(T,F,F,F)] # Our logical vector is 4. Returns every FOURTH element.
```
`R` knows to **recycle**.

Indexing Matrices and Data frames
====

Indexing matrices and data frames
----

Indexing vectors is a lot like indexing all other R objects.

  - You still use `[`
  - You can access values by name, index, or logicals

One difference:
  
  - Use the format  `x[a,b]` 
      + where `x` is a matrix or data frame object, 
      + `a` is the *row* index, 
      + `b` is the *column* index.

Indexing Matrices by **INDEX**
----

Let's work with the `volcano` data set.

```{r}
data(volcano)
dim(volcano)
class(volcano)
```

Indexing Matrices by **INDEX**
----

Use the `matrix[row, column]` format:
```{r}
volcano[1:6, 1:10] # show the first 6 rows and first 10 columns
```

Indexing Matrices by **INDEX**
----

Since we know we have `r nrow(volcano)` rows and `r ncol(volcano)` columns we know what possible values we can use to index `volcano`.

```{r}
volcano[52:58, 20:25]
```

Notice what R does when we don't have row names.

Indexing Matrices by **INDEX**
----

```{r}
# Returns specified rows within ALL COLUMNS
volcano[52:58,] # Leave out index for columns
```

Indexing Matrices by **INDEX**
----

```{r}
# Returns specified columns within ALL ROWS
volcano[,20:25] # Leave out index for columns
```

Indexing Matrices by **NAME**
----

```{r}
mat<-matrix(1:4, 2)
colnames(mat)<-c("A", "B")
rownames(mat)<-c("X", "Y")
mat
```

```{r}
mat["Y",]
```

Indexing dataframes
----

Let's look at the `dune.env` data set.

```{r warning=FALSE, message=FALSE, eval=-1}
install.packages("vegan") #IF you don't have it already
library(vegan)
```

```{r}
data(dune.env)
#head(dune.env, 4) # Can also use head with str()
str(dune.env)
```

Indexing dataframes by row
----

There's a huge difference between accessing a data frame by row *index* vs *row name*.

```{r}
dune.env[c(1,2,3),] # Access by INDEX
dune.env[c("1", "2", "3"),] # Access by ROWNAME
```

Indexing dataframes by **INDEX**
----

```{r}
data(varespec)
varespec[1:10, 1:5] # Taking a look at the upper left chunk of data
```

Indexing dataframes by **INDEX**
----

```{r}
varespec[c(1,2,3), 1:8] # Access by INDEX. (Same as varespec[1:3, 1:8])
```

Indexing dataframes by **ROWNAME**
----

```{r}
varespec[c("1", "2", "3"), 1:8] # Access by ROWNAME
rownames(varespec) # "1" is not a valid row name in `varespec`
```

This should be a **warning** to you about what to assign row names to avoid potentially big errors. When possible, assign descriptive, unique, short row names. Often times this is a sample ID.

Remember the `Birds` data frame?
----

```{r echo=-1}
source("Birds.R")
Birds
```

Each rowname is a unique, descriptive character identifier.

Indexing dataframes by column
----

You *can* index **columns** by column index, but ...

```{r}
dune.env[1:3, 3:4]
```

... it's preferable to use column names when possible---more clear, more **_explicit_**.

```{r}
dune.env[1:3, c("Management", "Use")]
```

Accessing whole vectors within a data frame
----

Use the `$`. After this the result can be worked with just like a vector by appending arguments within `[`.

```{r}
dune.env$A1
dune.env$A1[1:5]
```
In this form, you have to use the column name.

Accessing whole vectors within a data frame
----

Use the `$`. After this the result can be worked with just like a vector by appending arguments within `[`.

```{r}
dune.env$A1[1:5]
```

```{r}
dune.env[1:5,"A1"]
```

----

Using the `varespec` dataset, find the difference in cover values between 'Vaccviti' and 'Cladrang' at each site.

----

Using the `varespec` dataset, find the difference in cover values between 'Vaccviti' and 'Cladrang' at each site.

```{r}
varespec$Vaccviti-varespec$Cladrang
```

OR

```{r}
varespec[,"Vaccviti"]-varespec[,"Cladrang"]
```

----

Next, find the difference in cover values between 'Vaccviti' and 'Cladrang', but only in sites where the humus layer is less than 2. You will have to use the `varechem` dataset whose rownames ('sites') are the same as `varespec`.

----

Next, find the difference in cover values between 'Vaccviti' and 'Cladrang', but only in sites where the humus layer is less than 2. You will have to use the `varechem` dataset whose rownames ('sites') are the same as `varespec`.

```{r}
data(varechem)
varespec[varechem$Humdepth <2,"Vaccviti"]-varespec[varechem$Humdepth <2,"Cladrang"]
#Note we are using the data in one data frame to index another data frame.
```

Indexing Lists
====

Indexing lists
----

```{r}
data(dune.env)
```

Let's make a new list.
```{r}
dune.env$A1
fivenum(dune.env$A1) # See ?fivenum
```

Indexing lists
----

```{r}
dune.env$Management
table(dune.env$Management) # See ?table
```

Indexing lists
----

You can access the elements of a list in very similar ways. Let's make a list first.
```{r}
dune.summ<-list(DATA=dune.env, 
            SUMM.A1=fivenum(dune.env$A1), 
            TAB.MANAGE=table(dune.env$Management))
str(dune.summ)
```

Access list by name
----

```{r}
names(dune.summ)
dune.summ$SUMM.A1 # result of `fivenum(dune.env$A1)`
dune.summ$TAB.MANAGE # result of `table(dune.env$Management)`
```

Indexing Lists
----

We can also access a *copy* of elements within a list by item index.
```{r}
dune.summ[2] # Access a copy of the second element in dune.summ
```

 - produces the item *and its name*
 
Indexing Lists
----

More useful is to access the *actual* elements. What if we want to access the median (the 3rd element) in SUMM.A1? Three main ways:

```{r}
dune.summ$SUMM.A1[3]
```

Indexing Lists: the infamous `[[`
----

Notice the big difference between these two:
```{r}
dune.summ[2] # the element of the list
dune.summ[[2]] # the actual vector
```

**Double brackets allow access to the vector itself**, not just the element of the list.

Indexing Lists: the infamous `[[`
----

```{r }
dune.summ[2][3]
dune.summ[[2]][3]
```

Indexing lists
----

What if we want to access the **top 5 rows** in `DATA`? You try...

```{r}
str(dune.summ)
```

Indexing lists
----

What if we want to access the **top 5 rows** in `DATA`? First we need access to the actual data frame. After that, it's just like working with a normal data frame.

```{r}
dune.summ[["DATA"]][1:5,]
```
