---
title: "`ddply()` and Dates"
date: March 25, 2026
output: 
  ioslides_presentation:
    fig_width: 7
    fig_height: 6
    fig_caption: true
    widescreen: true
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "##")
```

Today
----

 - `plyr::summarize()` and `ddply()`
 - class `Date` and date arithmetic
 - **Combining the two**: splitting and summarizing data by date-derived groups
 - Practice with Archer Creek Watershed data


The limitation of `aggregate()`
----

You already know `aggregate()`:

```{r}
data(airquality)
aggregate(Temp ~ Month, airquality, FUN = mean)
```

But `aggregate()` only accepts **one** aggregation function at a time.

```{r, error=TRUE}
aggregate(Temp ~ Month, airquality, FUN = c(mean, sd))
```

`plyr::summarize()`
----

```{r}
library(plyr)
```

`summarize()` creates a new data frame from summary calculations. Unlike `aggregate()`, you can apply **multiple** functions at once.

```{r}
summarize(airquality, 
  Avg = mean(Temp), 
  StdDev = sd(Temp),
  Med = median(Temp))
```

 - You can name the output columns whatever you want

`plyr::summarize()`
----

Applied functions can reference objects created earlier in the same call:

```{r}
summarize(airquality, 
  Avg = mean(Temp), 
  Med = median(Temp), 
  Dif = abs(Avg - Med))
```

 - What capability are we missing?

`ddply()`
----

 - Also in the `plyr` package
 - "Split data frame, apply function, and return results in a data frame." --- `?ddply`
 - Hadley Wickham (2011). "The Split-Apply-Combine Strategy for Data Analysis", *Journal of Statistical Software, 40(1), 1-29*
 - **d**ataframe *in*, **d**ataframe *out*, ap**ply**

`ddply()` arguments
----

 - **`.data`** = data frame to be processed
 - **`.variables`** = variables to split data frame by
    + a character vector
    + as `.()`-quoted variables
    + a formula
 - **`.fun`** = function to apply to each piece
 - **`...`** = arguments passed to `.fun` (e.g., `na.rm`)
    + in our case `.fun` is `summarize()`

`ddply()` in action
----

```{r}
ddply(airquality, "Month", summarize,
  Avg.Temp = mean(Temp),
  Max.Ozone = max(Ozone, na.rm = TRUE))
```

 - Splits `airquality` by `Month`, applies `summarize()` to each piece, returns one data frame

`ddply()`: three ways to specify groups
----

```{r, eval=FALSE}
ddply(airquality, "Month", ...)          # character
ddply(airquality, .(Month), ...)         # .() notation
ddply(airquality, ~Month, ...)           # formula
```

All three produce identical results.

 - **`.`** is a function that quotes variable names without evaluating them (see `?.`)
 - The formula interface uses `~` just like `aggregate()`

`ddply()`: multiple grouping variables
----

```{r}
data(warpbreaks)
ddply(warpbreaks, .(wool, tension), summarize,
  Avg = mean(breaks),
  Med = median(breaks),
  Dif = abs(Avg - Med))
```

class `Date`
====

class `Date`
----

 - Looks like a character; is *not* just a character
 - Dates are represented *internally* as the number of days since 1970-01-01
    + Earlier dates are assigned negative values
 - Many methods work only with `Date` objects (arithmetic, sorting, plotting)
 - No times in class `Date` (see `Date-Time` classes for that)
    + Highest level of resolution is a day

Conversion to class `Date`
----

```{r}
x <- "03/25/2026"
class(x)
```

```{r}
d <- as.Date(x, format = "%m/%d/%Y")
class(d)
d
```

 - `?strptime` for the full list of format codes

Common date format codes
----

```{r, eval=FALSE}
?strptime
```

 - **%m** = month as decimal (01--12)
 - **%d** = day of month (01--31)
 - **%Y** = 4-digit year
 - **%y** = 2-digit year
 - **%j** = day of year (001--366)
 - **%A** = full weekday name
 - **%B** = full month name
 - **%b** = abbreviated month name

Alternate date notation
----

The separator and order don't matter --- just match the `format` argument to the input:

```{r}
as.Date("03/25/2026", format = "%m/%d/%Y")
as.Date("25-03-2026", format = "%d-%m-%Y")
as.Date("2026|03|25", format = "%Y|%m|%d")
```

 - The only *super* special character is **`%`**

Long-hand example
----

```{r}
Echar <- "Wednesday March 25, 2026"
Edate <- as.Date(Echar, format = "%A %B %d, %Y")
Edate
```

 - Enter separators (commas, spaces) verbatim in the format string

Conversion back to characters: `format()`
----

```{r}
Edate <- as.Date("2026-03-25")
format(Edate, "%d-%m-%Y")
format(Edate, "%B %Y")
format(Edate, "This is day %j of %Y.")
```

 - `format()` always returns a **character**, not a Date

Date arithmetic
----

Dates support arithmetic directly.

```{r}
d1 <- as.Date("2014-10-01")
d2 <- as.Date("2026-03-25")
d2 - d1
```

```{r}
d1 + 30
```

 - Subtraction returns a `difftime` object
 - Addition/subtraction of integers adds/removes days

`difftime()`
----

For more control over units, use `difftime()`:

```{r}
ss  <- as.Date("1962-09-27")  # Silent Spring published
ddt <- as.Date("1973-12-31")  # US DDT ban
difftime(ddt, ss, units = "weeks")
```

 - `units` accepts `"secs"`, `"mins"`, `"hours"`, `"days"`, `"weeks"`

Extracting info from Dates
----

```{r}
d <- as.Date("2026-03-25")
weekdays(d)
months(d)
quarters(d)
```

```{r}
format(d, "%j")  # Julian day
```

 - These functions return *characters*, not Dates

Extracting year and month with `format()`
----

```{r}
format(d, "%Y")       # year as character
as.numeric(format(d, "%m"))  # month as number
```


Building a date column
----

The `airquality` dataset has `Month` and `Day` columns but no actual `Date` column. Let's build one.

```{r}
# All observations are from 1973
airquality$Date <- as.Date(paste("1973", 
  airquality$Month, airquality$Day, sep = "-"))
head(airquality[, c("Month", "Day", "Date")])
```

 - `paste()` builds the date string; `as.Date()` parses it
 - Default format is `"%Y-%m-%d"` so no `format` argument needed

Why bother with a Date column?
----

Once you have a real `Date`, you can derive grouping variables you never had:

```{r}
airquality$Weekday <- weekdays(airquality$Date)
airquality$Quarter <- quarters(airquality$Date)
airquality$Week    <- as.numeric(format(airquality$Date, "%U"))
```

Now we can split by weekday, quarter, or week number --- none of which existed in the raw data.

`ddply()` by weekday
----

Do ozone levels differ by day of the week? (Hint: industrial emissions tend to drop on weekends.)

```{r}
ddply(airquality, "Weekday", summarize,
  N = sum(!is.na(Ozone)),
  Avg.Ozone = mean(Ozone, na.rm = TRUE),
  Avg.Temp  = mean(Temp))
```

`ddply()` by weekday
----

The output is alphabetical. To order by day of week:

```{r}
airquality$Weekday <- factor(airquality$Weekday,
  levels = c("Sunday","Monday","Tuesday","Wednesday",
             "Thursday","Friday","Saturday"),
  ordered = TRUE)

ozone_by_day <- ddply(airquality, "Weekday", summarize,
  N = sum(!is.na(Ozone)),
  Avg.Ozone = round(mean(Ozone, na.rm = TRUE), 1))
ozone_by_day
```

`ddply()` by week number
----

Summarize temperature by week of the year:

```{r}
temp_by_week <- ddply(airquality, "Week", summarize,
  N = length(Temp),
  Avg.Temp = mean(Temp),
  Range = max(Temp) - min(Temp))
head(temp_by_week, 8)
```

Plotting weekly summaries
----

```{r, fig.height=4}
plot(Avg.Temp ~ Week, data = temp_by_week, type = "b",
  pch = 19, col = "darkred",
  ylab = expression("Mean Temperature ("*degree*"F)"),
  xlab = "Week of Year",
  main = "NYC Weekly Temperature, 1973")
```

 - `ddply()` produces a clean data frame ready for plotting

Multiple grouping variables
----

Split by **both** month and whether it is a weekend:

```{r}
airquality$Weekend <- airquality$Weekday %in% 
  c("Saturday", "Sunday")

ddply(airquality, .(Month, Weekend), summarize,
  N = sum(!is.na(Ozone)),
  Avg.Ozone = round(mean(Ozone, na.rm = TRUE), 1))
```

Using `ifelse()` to create custom groups
----

What if we want to group by the first vs. second half of each month?

```{r}
airquality$Half <- ifelse(airquality$Day <= 15, 
  "1st Half", "2nd Half")

ddply(airquality, .(Month, Half), summarize,
  Avg.Wind = round(mean(Wind), 1),
  Avg.Temp = round(mean(Temp), 1))
```

 - `ifelse()` creates the grouping variable; `ddply()` does the work

Practical example: Ackerman Clearing weather station
====

The data
----

Ackerman Clearing is a meteorological station in the Huntington Wildlife Forest, Adirondack Park. Daily observations include air temperature, precipitation, snow depth, and wind speed.

```{r}
acw <- read.csv("ACW-met.csv")
str(acw)
```

First look
----

```{r}
head(acw[, c("timestamp","air_temp_avg","rain",
  "snow_depth_mean","windspeed_avg")])
```

 - `timestamp` is a character --- we need to convert it to a `DateTime` 

You try...
----

```{r eval=FALSE}
head(acw)
?strptime
?DateTimeClasses
```

Converting the timestamp
----

```{r}
class(acw$timestamp)
acw$Date <- as.Date(acw$timestamp, 
  format = "%Y-%m-%d %H:%M:%S")
class(acw$Date)
```

```{r}
range(acw$Date)
```

 - About 3 years of daily data (2018--2020)

Deriving grouping variables from the Date
----

```{r}
acw$Year    <- as.numeric(format(acw$Date, "%Y"))
acw$Month   <- as.numeric(format(acw$Date, "%m"))
acw$DOY     <- as.numeric(format(acw$Date, "%j"))
acw$Weekday <- weekdays(acw$Date)
```

```{r}
head(acw[, c("Date","Year","Month","DOY")], 4)
```

Exercise
----

Use `ddply()` to give you mean, min, and max monthly temperatures by `Year`.

Monthly temperature by year
----

```{r}
ddply(acw, .(Year, Month), plyr::summarize,
  Mean.Temp = round(mean(air_temp_avg), 1),
  Min.Temp  = round(min(air_temp_min), 1),
  Max.Temp  = round(max(air_temp_max), 1))
```


Plotting monthly means
----

```{r, fig.height=4, eval=F}
monthly <- ddply(acw, .(Year, Month), plyr::summarize,
  Mean.Temp = mean(air_temp_avg))

plot(Mean.Temp ~ Month, data = monthly, 
  col = as.factor(monthly$Year), pch = 19,
  ylab = expression("Mean Temperature ("*degree*"C)"),
  main = "Ackerman Clearing Monthly Temps")
legend("topleft", legend = unique(monthly$Year),
  col = 1:3, pch = 19, bty = "n")
```

Plotting monthly means
----

```{r, fig.height=4, eval=T, echo=F}
monthly <- ddply(acw, .(Year, Month), plyr::summarize,
  Mean.Temp = mean(air_temp_avg))

plot(Mean.Temp ~ Month, data = monthly, 
  col = as.factor(monthly$Year), pch = 19,
  ylab = expression("Mean Temperature ("*degree*"C)"),
  main = "Ackerman Clearing Monthly Temps")
legend("topleft", legend = unique(monthly$Year),
  col = 1:3, pch = 19, bty = "n")
```


`ddply()` vs. `dplyr`
----

`plyr` pioneered split-apply-combine in R. `dplyr` (also by Wickham) is its successor, built for speed and pipelines.

The core translation:

```{r, eval=FALSE}
# plyr
ddply(df, .(var1, var2), summarize, ...)

# dplyr
df |> group_by(var1, var2) |> summarise(...)
```

 - `group_by()` replaces the `.variables` argument
 - The pipe (`|>`) replaces nesting
 - `summarise()` replaces both `.fun` and `...`

Side-by-side: simple grouping
----

```{r, message=FALSE}
library(dplyr)
```

```{r}
# plyr
ddply(acw, .(Year, Month), plyr::summarize,
  Mean.Temp = round(mean(air_temp_avg), 1),
  Min.Temp  = round(min(air_temp_min), 1),
  Max.Temp  = round(max(air_temp_max), 1))
```

Side-by-side: simple grouping
----

```{r}
# dplyr

acw |>
  group_by(Year, Month) |>
  summarise(
    Mean.Temp = round(mean(air_temp_avg), 1),
    Min.Temp  = round(min(air_temp_min), 1),
    Max.Temp  = round(max(air_temp_max), 1),
    .groups = "drop") #Drops persistent grouping. See `?dplyr_by`.
```

 - Identical results; `dplyr` returns a `tibble` (prints slightly differently)

References
----

 - `?ddply`, `?summarize` (package `plyr`)
 - `?as.Date`, `?strptime`, `?difftime`, `?seq.Date`
 - Wickham, H. (2011). The Split-Apply-Combine Strategy for Data Analysis. *Journal of Statistical Software*, 40(1), 1--29.
 - Ackerman Clearing meteorological data, Huntington Wildlife Forest, Adirondack Park (adk-ltm.org)
 - `?airquality`
