---
title: "Summarizing with `ddply()`"
author: "Hyatt Green"
date: "March 18, 2026"
output: 
  ioslides_presentation:
    fig_width: 7
    fig_height: 6
    fig_caption: true
    widescreen: true
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Today

 - Review `aggregate()`
 - `plyr::summarize()`
 - **`ddply()`**
 - The `d_ply()` family

## Remember `aggregate()`?

```{r}
data(warpbreaks)
aggregate(breaks~wool, warpbreaks, FUN=mean)
```

## Remember `aggregate()`?

```{r}
aggregate(breaks~wool+tension, warpbreaks, FUN=mean)
```

## `aggregate()` limitation

Only one aggregation function allowed.

```{r, error=TRUE}
# Does not work
aggregate(breaks~wool+tension, warpbreaks, FUN=c(mean, sd))
```

 - What if we want the mean **and** the sd **and** the median in one call?

## `plyr::summarize()`

 - Creates new dataframe

```{r, message=FALSE}
library(plyr)
summarize(warpbreaks, mean(breaks), sd(breaks))
```

## `plyr::summarize()`

 - Usually add custom names

```{r}
summarize(warpbreaks, 
  Avg = mean(breaks), 
  StdDev = sd(breaks))
```

## `plyr::summarize()`

 - Applied functions can be complex

```{r}
summarize(warpbreaks, 
  Avg = mean(breaks), 
  Med = median(breaks), 
  Dif = abs(Avg - Med)) #Can use objects created within function
```

 - What capability are we missing here with `summarize()`?

## `ddply()`

 - Also in the `plyr` package
 - "Split data frame, apply function, and return results in a data frame." --- ?ddply
 - Hadley Wickham (2011). "The Split-Apply-Combine Strategy for Data Analysis", *Journal of Statistical Software, 40(1), 1-29*
 - **d**ataframe *in*, **d**ataframe *out*, ap**ply**

## `ddply()` arguments

`ddply()` arguments:
 
 - **`.data`** = data frame to be processed
 - **`.variables`** = variables to split data frame by
    + a character vector
    + as `as.quoted()` variables
    + a formula
 - **`.fun`** = function to apply to each piece
 - **`...`** = arguments passed to **`.fun`** (e.g., `na.rm`)
    + in our case `.fun` is `summarize()`

## `ddply()` in action

 - Often used with `summarize()` to apply more than one function

```{r}
ddply(warpbreaks, "wool", summarize,
   Avg = mean(breaks),
   Med = median(breaks),
   Dif = abs(Avg - Med))
```

## `ddply()`: `.()` notation

```{r}
 ddply(warpbreaks, .(wool, tension), summarize,
   Avg = mean(breaks),
   Med = median(breaks),
   Dif = abs(Avg - Med))
```

 - **`.`** is a function that simply quotes the names of each variable (see `?.`)
      + does not evaluate the variable
      + simply allows you to drop quotes

## `ddply()`: formula interface

```{r}
 ddply(warpbreaks, ~wool+tension, summarize,
   Avg = mean(breaks),
   Med = median(breaks),
   Dif = abs(Avg - Med))
```


## Three ways to specify variables

| Method | Syntax | Notes |
|---|---|---|
| Character vector | `"wool"` | Quoted, single variable |
| `.()` notation | `.(wool, tension)` | Unquoted, multiple variables |
| Formula | `~wool+tension` | Familiar from `aggregate()` |

All three produce the same result. Use whichever you find most readable.


## The `d_ply()` family

What if we want to output a list? Use `dlply()`...

```{r}
dlply(warpbreaks, .(wool, tension), summarize,
   Avg = mean(breaks),
   Med = median(breaks),
   Dif = abs(Avg - Med))
```

## The `d_ply()` family

What if we want to output an array? Use `daply()`...

```{r}
daply(warpbreaks, .(wool, tension), summarize,
   Avg = mean(breaks),
   Med = median(breaks),
   Dif = abs(Avg - Med))
```

## The `d_ply()` family

By far, `ddply()` is most common because dataframes are easiest to work with.

Other functions that *produce* dataframes:

 - `ldply()`
 - `adply()`
 - `mdply()`


## Practice with `ddply()`

Import the `birthwt` data set from the **MASS** package. For *all combinations of race and smoking status*, find the:

 - number of observations within each group
 - mean and std. dev. of:
    + mother's weight at last menstrual period
    + birth weight
 - median number of physician visits during the first trimester

----

```{r, message=FALSE}
library(MASS)
data(birthwt)
str(birthwt)
```

----

```{r}
head(birthwt)
```

----

```{r}
ddply(birthwt, .(race, smoke), summarize, 
  N = length(bwt),
  MW.mean = mean(lwt),
  MW.sd = sd(lwt),
  BWT.mean = mean(bwt),
  BWT.sd = sd(bwt),
  FTV.med = median(ftv) )
```

----

Wrap in `round()` for cleaner output:

```{r}
round(ddply(birthwt, .(race, smoke), summarize,
  N = length(bwt),
  MW.mean = mean(lwt),
  MW.sd = sd(lwt),
  BWT.mean = mean(bwt),
  BWT.sd = sd(bwt),
  FTV.med = median(ftv) ) )
```


## `aggregate()` vs `ddply()`

| Feature | `aggregate()` | `ddply()` |
|---|---|---|
| Package | base R | `plyr` |
| Functions per call | **One** | **Many** |
| Variable specification | Formula (`y ~ x`) | Character, `.()`, or formula |
| Output | Data frame | Data frame (also list, array via `dlply`, `daply`) |
| Custom column names | No (uses function name) | Yes (name each summary) |
| Cross-referencing summaries | No | Yes (`Dif = abs(Avg - Med)`) |

 > - Use `aggregate()` for quick one-function summaries
 > - Use `ddply()` when you need multiple summaries in one table

## Summary

 > - `plyr::summarize()` applies multiple functions but has no grouping
 > - `ddply()` = `summarize()` + grouping = the full toolkit
 > - **d**ataframe in, **d**ataframe out, ap**ply**
 > - Three equivalent ways to specify grouping variables
 > - The `d_ply()` family: `ddply`, `dlply`, `daply` for different output types
