---
title: "Homework 7: Writing and Debugging Functions"
author: 
date: "Assigned Mar 4, 2026; Due Mar 18, 2026"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions

Complete all exercises below by filling in your answers in the designated spaces. When finished:

1. Use *this* .Rmd file to complete the homework
2. Complete the homework creating code chunks where needed
3. Knit the document to HTML (click the "Knit" button or press Ctrl+Shift+K / Cmd+Shift+K)
4. Submit **both** your .Rmd file and the knitted Word/PDF (HTMLs are not 'uploadable', but can be converted to PDF) to Blackboard

**Due:** Wednesday, March 18 by 11:59 PM

---

## Part 1: Building Your First Functions

### Question 1 (1 point)

Foresters use **diameter at breast height** (DBH, in cm) to estimate a tree's basal area --- the cross-sectional area of the trunk. The formula is:

$$BA = \frac{\pi}{4} \times DBH^2$$

where $BA$ is in cm$^2$.

a) Write a function called `basal_area` that takes a single argument `dbh` (in cm) and returns the basal area. Give `dbh` a default value of 25.

b) Test your function on these inputs and show the output:
   - `basal_area()` (using the default)
   - `basal_area(10)`
   - `basal_area(c(10, 25, 40, 55))`

c) Notice that your function works on a single number *and* on a vector. In one sentence, explain why this happens without you having to write any special loop or extra code.

*Delete this text and write your explanation here.*

---

### Question 2 (1 point)

Wildlife biologists sometimes need to convert animal body mass between units depending on the journal, the country, or the dataset they inherited. Writing a converter saves time and prevents arithmetic mistakes.

a) Write a function called `mass_convert` that takes three arguments:

   - `mass` --- a numeric vector (no default; the user must supply this)
   - `from` --- a character string, with a default of `"kg"`
   - `to` --- a character string, with a default of `"lb"`

   The function should handle conversions between `"kg"`, `"lb"`, and `"g"` (grams). Use the conversion factors: 1 kg = 2.20462 lb = 1000 g. If `from` and `to` are the same, just return the original mass.

   *Hint:* one clean approach is to first convert everything to kg (regardless of `from`), then convert from kg to the target unit.

b) Show the output of each of the following:

```{r q2-demo, eval=FALSE}
mass_convert(100)                          # 100 kg to lb
mass_convert(100, from = "lb", to = "kg")  # 100 lb to kg
mass_convert(5000, from = "g", to = "lb")  # 5000 g to lb
mass_convert(80, from = "kg", to = "kg")   # no conversion needed
```

c) The average mass of an adult male black bear in the Adirondacks is roughly 125 kg. Use your function to express that in pounds and in grams.

*Delete this text and show your code here.*

---

## Part 2: Functions That Summarize Data

### Question 3 (1.5 points)

Aquatic ecologists assess water quality partly by looking at dissolved oxygen (DO) readings from a stream over time. A quick health check might report the mean, the minimum (stressful for fish), and how many readings fall below a critical threshold.

a) Write a function called `do_report` that takes two arguments:

   - `do_vec` --- a numeric vector of dissolved oxygen readings (mg/L)
   - `threshold` --- a numeric value representing the stress threshold (default = 5 mg/L)

   The function should return a **named list** with four elements:

   - `mean_do` --- the mean of `do_vec` (handle `NA`s)
   - `min_do` --- the minimum value (handle `NA`s)
   - `n_below` --- the count of readings below the threshold (ignore `NA`s)
   - `pct_below` --- the percentage of readings below the threshold, rounded to one decimal place (ignore `NA`s)

b) Create a test vector of dissolved oxygen readings to simulate a warm summer month on a sluggish creek:

```{r q3-data, eval=FALSE}
creek_do <- c(6.2, 5.8, 4.1, 3.9, 7.0, 5.5, 4.8, 6.1, NA, 3.2, 
              5.0, 4.5, 6.8, 5.3, 4.0, 7.2, 5.9, 3.5, 4.7, 6.0)
```

   Run `do_report(creek_do)` and show the output.

c) Now run `do_report(creek_do, threshold = 4.5)`. How does raising the bar change the picture?

d) Your function returns a list. Show how you would extract *just* the `pct_below` element from the result using `$` notation.

*Delete this text and write your explanation here.*

---

### Question 4 (1 point)

The `trees` dataset built into R contains measurements of 31 black cherry trees: `Girth` (diameter in inches), `Height` (in feet), and `Volume` (in cubic feet). Foresters estimate timber volume from these measurements using various allometric equations. A simple one is:

$$V = a \times G^b \times H^c$$

where $G$ is girth, $H$ is height, and $a$, $b$, $c$ are fitted constants.

a) Write a function called `est_volume` that takes five arguments: `girth`, `height`, `a`, `b`, and `c`. Give the constants these defaults: `a = 0.003`, `b = 2.0`, `c = 1.0`. The function should return the estimated volume.

b) Load the `trees` dataset. Use your function to estimate volume for every tree using the defaults:

```{r q4-demo, eval=FALSE}
data(trees)
v_est <- est_volume(trees$Girth, trees$Height)
```

c) Now compare your estimates to the actual `Volume` column. Compute the **mean absolute error**:

$$MAE = \frac{1}{n}\sum_{i=1}^{n}|V_{estimated,i} - V_{actual,i}|$$

   You can do this in one line using `mean()` and `abs()`.

d) Try tweaking the constants to reduce the MAE. Report the values of `a`, `b`, and `c` you settled on and the resulting MAE. You don't need to find the best possible fit --- just show that changing the arguments changes the output.

*Delete this text and write your explanation here.*

---

## Part 3: Scope and Debugging

### Question 5 (0.75 points)

a) Run the following code exactly as written and explain the output of each `exists()` call:

```{r q5-demo, eval=FALSE}
wind_chill <- function(temp_f, wind_mph) {
  wc <- 35.74 + 0.6215 * temp_f - 35.75 * wind_mph^0.16 + 
        0.4275 * temp_f * wind_mph^0.16
  return(wc)
}

wind_chill(20, 15)
exists("wc")
exists("wind_chill")
exists("temp_f")
```

b) In your own words, explain why `wc` and `temp_f` do not exist in your global environment after you call the function, even though both were clearly created or used inside it.

c) Suppose you wanted to store the wind chill for a 20-degree day with 15 mph winds so you could use it later in your script. Write the one line of code that accomplishes this.

*Delete this text and write your explanation here.*

---

### Question 6 (0.75 points)

A colleague in the wildlife department wrote a function to classify black bear body condition from mass (in kg), but it's not working. Here is their code:

```{r q6-buggy, eval=FALSE}
bear_condition <- function(mass_kg) {
  if (mass_kg < 60) {
    condition <- "underweight"
  } else if (mass_kg >= 60 & mass_kg <= 100) {
    condition <- "normal"
  } else if (mass_kg > 100) {
    condtion <- "heavy"
  }
  return(condition)
}
```

a) Test the function with `bear_condition(50)` and `bear_condition(80)`. These work fine. Now test it with `bear_condition(130)`. What happens?

b) Identify the bug. Describe the problem in one sentence.

c) Fix the function and show that it works.

d) Add `browser()` as the first line inside the function. Run `bear_condition(130)` again. In the browser console, step through the code line by line (type `n` at the prompt). At each step, check whether `condition` exists yet by typing `condition`. At what point does the problem become visible? Type `Q` to quit the browser. Remove `browser()` when finished.

Explain in your own words how `browser()` helped you find the issue:

*Delete this text and write your explanation here.*

---

## Submission Checklist

Before submitting, make sure you have:

- [ ] Filled in your name in the author field at the top
- [ ] Answered all questions
- [ ] Written explanations in your own words where requested
- [ ] Successfully knitted your document to HTML (or Word/PDF)
- [ ] Uploaded both your .Rmd file AND your knitted output to Blackboard

---

*This homework is worth 6 points total.*
