---
title: "Merging Data"
date: "Mar 2, 2026"
output: 
  ioslides_presentation:
    fig_width: 7
    fig_height: 6
    fig_caption: true
    widescreen: true
    css: style.css
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Today

 - Why merge?
 - `merge()` and its arguments
 - The four types of merge
 - Merging by differently named columns
 - The `suffixes` argument
 - Many-to-one merges
 - Common pitfalls and debugging
 - Exercises

## Why merge?

Real data almost never arrives in one tidy table.

 - Field measurements in one file, lab results in another
 - Species observations in one table, trait data in another
 - One sensor logs temperature, another logs precipitation
 
The problem: **how do you combine two data frames that share some---but not all---rows and columns?**

## The three questions

Before you type `merge()`, answer three questions:

 >1. **Which** two data frames am I combining?
 2. **By which column(s)** should they be matched?
 3. **What happens to rows** that exist in one frame but not the other?

## `merge()` {.smaller}

```{r, eval=FALSE}
?merge
```

<PRE>
merge(x, y, by = intersect(names(x), names(y)),
      by.x = by, by.y = by, all = FALSE, all.x = all, all.y = all,
      sort = TRUE, suffixes = c(".x",".y"),
      incomparables = NULL, ...)
</PRE>

| Argument | What it does |
|----------|--------------|
| `x`, `y` | The two data frames |
| `by` | Column(s) to match on (name or index) |
| `by.x`, `by.y` | Use when column names differ between frames |
| `all` | Keep all rows from *both* frames? |
| `all.x` | Keep all rows from `x`? |
| `all.y` | Keep all rows from `y`? |
| `suffixes` | Disambiguate duplicate column names |

## Example data: water quality

```{r}
wq <- data.frame(
  month = c("Aug", "Sept", "Oct", "Nov"),
  DO    = c(12.3, 17.2, 14.0, 8.9),
  pH    = c(7.2, 6.9, 7.8, 7.7)
)
ec <- data.frame(
  month = c("Sept", "Oct", "Nov", "Dec"),
  Ecoli = c(120, 2080, 830, 7900)
)
```

 - Both have `month`
 - `wq` has Aug but not Dec
 - `ec` has Dec but not Aug

## Check names before merging

```{r}
intersect(names(wq), names(ec))
```

This is what `merge()` uses by default for `by`. Always check this yourself first, especially with large data frames where columns may share names by accident.

## Inner merge (the default)

```{r}
merge(wq, ec)
```

 - Only rows found in **both** frames are kept
 - `Aug` dropped (not in `ec`); `Dec` dropped (not in `wq`)
 - Result sorted alphabetically by `month`

This is the database equivalent of an **inner join**.

## Outer merge: `all = TRUE`

```{r}
merge(wq, ec, all = TRUE)
```

 - **All** rows from **both** frames kept
 - Missing values filled with `NA`
 - This is a **full outer join**

## Left merge: `all.x = TRUE`

```{r}
merge(wq, ec, all.x = TRUE)
```

 - All rows from `x` (`wq`) kept
 - Unmatched rows from `y` (`ec`) dropped
 - `Aug` stays; `Dec` goes
 - This is a **left join**

## Right merge: `all.y = TRUE`

```{r}
merge(wq, ec, all.y = TRUE)
```

 - All rows from `y` (`ec`) kept
 - Unmatched rows from `x` (`wq`) dropped
 - `Dec` stays; `Aug` goes
 - This is a **right join**

## The four merges, summarized {.smaller}

| Type | R code | Keeps from `x` | Keeps from `y` | SQL equivalent |
|------|--------|-----------------|-----------------|----------------|
| Inner | `merge(x, y)` | Matched only | Matched only | `INNER JOIN` |
| Full outer | `merge(x, y, all=T)` | All | All | `FULL OUTER JOIN` |
| Left | `merge(x, y, all.x=T)` | All | Matched only | `LEFT JOIN` |
| Right | `merge(x, y, all.y=T)` | Matched only | All | `RIGHT JOIN` |

**Rule of thumb:** if you're not sure, start with `all = TRUE` so you can *see* what didn't match, then decide what to discard.

## When column names don't match

Suppose the `ec` data used `sampling_month` instead of `month`:

```{r}
ec2 <- ec
names(ec2)[1] <- "sampling_month"
names(ec2)
```

## `by.x` and `by.y`

```{r}
merge(wq, ec2, by.x = "month", by.y = "sampling_month")
```

 - `by.x` says which column in `x` to match on
 - `by.y` says which column in `y` to match on
 - The result uses the name from `x`

## Merging by multiple columns

Sometimes a single column isn't enough to uniquely identify a row. Common in ecological data where you have site + year, or plot + date.

```{r}
sites_a <- data.frame(
  site = c("A", "A", "B", "B"),
  year = c(2020, 2021, 2020, 2021),
  temp = c(15.2, 16.1, 12.8, 13.5)
)
sites_b <- data.frame(
  site = c("A", "A", "B", "B"),
  year = c(2020, 2021, 2020, 2021),
  precip = c(820, 790, 1100, 1050)
)
```

## Merging by multiple columns

```{r}
merge(sites_a, sites_b)
```

 - `intersect(names(sites_a), names(sites_b))` returns `c("site", "year")`
 - Both columns used automatically
 - Each site-year combination matched correctly

## The `suffixes` argument

What happens when both data frames have a column with the same name that is *not* the merge key?

```{r}
df1 <- data.frame(id = 1:3, value = c(10, 20, 30))
df2 <- data.frame(id = 1:3, value = c(100, 200, 300))
merge(df1, df2)
```

## Customizing suffixes

The defaults `.x` and `.y` aren't very informative. Use `suffixes` to make them meaningful:

```{r}
merge(df1, df2, suffixes = c("_field", "_lab"))
```

This is much more readable when you come back to your code six months later.

## Many-to-one merges

Not all merges are one-to-one. Sometimes you want to attach a lookup table where one frame has fewer rows:

```{r}
observations <- data.frame(
  species = c("PIST", "PIST", "ACRU", "ACRU", "QURU"),
  dbh_cm  = c(25, 32, 18, 22, 45)
)
lookup <- data.frame(
  species   = c("PIST", "ACRU", "QURU"),
  common    = c("White Pine", "Red Maple", "Red Oak")
)
```

## Many-to-one merges

```{r}
merge(observations, lookup)
```

 - `lookup` has 3 rows, `observations` has 5
 - Each species code in `observations` matched to its common name
 - Rows in `observations` are *not* lost---they expand to include the lookup info
 - This is one of the most useful merge patterns

## Checking your merge with `dim()` {.smaller}

**Always check dimensions before and after merging.**

```{r}
dim(wq)
dim(ec)
dim(merge(wq, ec))           # inner
dim(merge(wq, ec, all = TRUE)) # full outer
```

If the merged frame is unexpectedly large, you likely have duplicate keys causing a **Cartesian product** (every row in `x` paired with every matching row in `y`).

## Common pitfall: accidental Cartesian product

```{r}
d1 <- data.frame(id = c(1, 1, 2), val1 = c("a", "b", "c"))
d2 <- data.frame(id = c(1, 1, 2), val2 = c("x", "y", "z"))
merge(d1, d2)
```

 - `id = 1` appears twice in **both** frames
 - Result: 2 x 2 = 4 rows for `id = 1`
 - This blows up fast with real data

## Debugging checklist

When `merge()` gives unexpected results:

 1. **Check `intersect(names(x), names(y))`** --- are the right columns being matched?
 2. **Check `dim()` before and after** --- did the row count change the way you expected?
 3. **Check for duplicate keys** --- `table(table(x[, by_cols]))` should show all 1s if keys are unique
 4. **Check spelling** --- `"Sept"` vs `"September"` will *not* match
 5. **Check class** --- merging a factor column with a character column can produce surprises

## Quick reference: `identical()`

`identical()` tests whether two objects are **exactly the same**:

```{r}
wq2 <- merge(wq, ec, by = "month")
wq3 <- merge(wq, ec, by = 1)
wq4 <- merge(wq, ec)
identical(wq2, wq3)
identical(wq3, wq4)
```

Useful for confirming that different `merge()` calls produce the same result.

## Tidyverse joins: `dplyr`

The `dplyr` package provides a family of `_join()` functions that do the same thing as `merge()` but with more explicit naming.

```{r, message=FALSE}
library(dplyr)
```

## `dplyr` join functions {.smaller}

Each join type gets its own function---no toggling `all`, `all.x`, `all.y`:

| `dplyr` function | Equivalent `merge()` | Keeps |
|------------------|----------------------|-------|
| `inner_join(x, y)` | `merge(x, y)` | Rows in **both** |
| `full_join(x, y)` | `merge(x, y, all = TRUE)` | **All** rows |
| `left_join(x, y)` | `merge(x, y, all.x = TRUE)` | All rows from `x` |
| `right_join(x, y)` | `merge(x, y, all.y = TRUE)` | All rows from `y` |

The `by` argument works similarly, and column name mismatches use `by = c("name_in_x" = "name_in_y")` syntax.

## `dplyr` join examples

```{r}
inner_join(wq, ec, by = "month")
```

```{r}
left_join(wq, ec, by = "month")
```

## Handling mismatched names in `dplyr`

Recall `ec2` where the column was named `sampling_month`:

```{r}
full_join(wq, ec2, by = c("month" = "sampling_month"))
```

This is the `dplyr` equivalent of `merge(wq, ec2, by.x = "month", by.y = "sampling_month", all = TRUE)`.

## Filtering joins in `dplyr`

`dplyr` also has two joins that don't add columns---they just filter rows:

```{r}
semi_join(wq, ec, by = "month")  # rows in wq that HAVE a match in ec
```

```{r}
anti_join(wq, ec, by = "month")  # rows in wq that have NO match in ec
```

These have no direct `merge()` equivalent and are genuinely useful for data QC---e.g., "which of my field sites are missing from the lab results?"

## `merge()` vs `dplyr` joins

**Advantages of `merge()`:**

 - No package dependency (base R)
 - One function to learn
 - Fine for most datasets

**Advantages of `dplyr` joins:**

 - Function names make the join type explicit
 - Preserves row order (base `merge()` sorts by default)
 - `semi_join()` and `anti_join()` are handy extras
 - Somewhat faster on large data
 

## Other alternatives: `data.table`

 - `data.table::merge()` uses the **same arguments** as base `merge()`, so switching is nearly free
 - Speed advantage is dramatic: merges on millions of rows that take minutes in base R finish in seconds
 - Memory-efficient---modifies data in place rather than copying
 - The package also has its own `[` syntax for joins (`X[Y, on = ...]`) which is extremely concise once learned
 - Worth adopting if your data routinely exceeds ~1 million rows or if you're doing repeated merges in a loop

## Other alternatives: `sqldf`

 - `sqldf::sqldf()` lets you write **actual SQL queries** directly on R data frames
 - Useful if you already know SQL from database work or GIS (e.g., PostGIS, Access, SQLite)
 - Handles complex joins, subqueries, and aggregations in a single statement
 - A good bridge if you're moving between R and database workflows

## Exercise 1

Create these two data frames and perform the four types of merge (inner, full outer, left, right). For each, check the dimensions and identify which rows were kept or dropped:

```{r, eval = FALSE}
growth <- data.frame(
  plot = c("P1", "P2", "P3", "P4"),
  height_cm = c(120, 95, 140, 110)
)
soil <- data.frame(
  plot = c("P2", "P3", "P4", "P5"),
  organic_pct = c(3.2, 5.1, 2.8, 4.5)
)
```


## Summary

 - `merge()` combines two data frames by shared columns
 - Four types: inner (default), full outer, left, right
 - Always check `intersect(names(x), names(y))` before merging
 - Always check `dim()` before and after
 - Use `by.x` and `by.y` when column names differ
 - Use `suffixes` to make duplicate column names informative
 - Watch out for Cartesian products from duplicate keys
 - Start with `all = TRUE` when exploring, then restrict as needed
